---
title: "GEO Normal Tissue Gene Expression/Variant Data"
author: "Dustin Miller"
date: "10/25/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, python.reticulate = FALSE)
#Load packag(es)
library(tidyverse)
library(magrittr)
library(stats)
```

### Before you begin
Please save this file to a directory where you want your data saved to. Then, set your working directory in RStudio to the source of the file location.

### Download and read in data from GEO that includes patient ID and cancer type. Series: [GSE62944](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE62944)
```{r}
geoNormalData <- read_tsv("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE62944&format=file&file=GSE62944%5F06%5F01%5F15%5FTCGA%5F24%5FNormal%5FCancerType%5FSamples%2Etxt%2Egz/GSE62944_06_01_15_TCGA_24_Normal_CancerType_Samples.txt.gz", col_names = c("patient", "cancerType"))

```

### Download tar file with GEO raw expression data. This tar file contains various txt files. Series: [GSE62944](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE62944)
```{r}
#Default will download to users downloads folder unless user changes "destfile = "
#download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE62944&format=file", destfile = "GSE62944_RAW.tar")
```

### Extract TPM file from tar file and remove tar file
```{bash}
#Assuming "destfile = " above was not changed, this will extract the TPM file of interest and remove files not of interest. If "destfile = " was changes agove, change cd directory accordingly.
#tar zxvf GSE62944_RAW.tar GSM1697009_06_01_15_TCGA_24.normal_Rsubread_TPM.txt.gz
#rm -r GSE62944_RAW.tar
```

### Read in TPM file to RStudio. This file contains gene expression values of normal tissues for each patient
```{r}
expressionDataNormal <- read_tsv("GSM1697009_06_01_15_TCGA_24.normal_Rsubread_TPM.txt.gz")
```

### Determine how many patients have particlar cancer type and view top 5
```{r}
cancerTypes <- count(geoNormalData, cancerType) %>% arrange(desc(n))
head(cancerTypes, 5)
```

### Create dataset with only BRCA data and another with KIRC as these cancer types have highest number of patients
```{r}
brcaData <- filter(geoNormalData, cancerType == 'BRCA')

kircData <- filter(geoNormalData, cancerType == 'KIRC')
```

### Transpose expressionData to place patients as rows and genes as columns 
```{r}
expressionDataTranspose <- column_to_rownames(expressionDataNormal, var = "X1") %>% t()
expressionDataTranspose <- cbind(patient = rownames(expressionDataTranspose), expressionDataTranspose)
```

### Join BRCA only patients and KIRC only patients to gene expression data, making totalData data frame for each cancer type
```{r}
brcaTotalData <- inner_join(as_tibble(expressionDataTranspose), brcaData, by = "patient")
kircTotalData <- inner_join(as_tibble(expressionDataTranspose), kircData, by = "patient")
```

### Download Teir 1 data only from [Cancer Gene Census](https://cancer.sanger.ac.uk/cosmic/census?tier=1) and place in current working directory. This was downloaded, Oct. 22, 2018.
```{r}
censusData <- read_tsv("Census_allTue Oct 23 01_01_01 2018.tsv")
```

### Keep only data where gene has been found to be mutated in germline and is a TSG-only
```{r}
censusTSG <- filter(censusData, censusData$`Role in Cancer` == "TSG", censusData$Germline == "yes")
vectorTSG <- as.vector(censusTSG$`Gene Symbol`)
```

### Create new datasets that include only TSG's from vectorTSG in brcaTotalData and kircTotalData
```{r}
brcaTotalDataTsg <- select(brcaTotalData, patient, vectorTSG)
kircTotalDataTsg <- select(kircTotalData, patient, vectorTSG)
write_tsv(brcaTotalDataTsg,"brcaTotalDataTsg.tsv")
write_tsv(kircTotalDataTsg,"kircTotalDataTsg.tsv")
```

### This python script is used to make the brcaTotalDataTsg dataset tidy and remove last portion of patient IDs
```{python}
import re
with open("brcaTotalDataTsg.tsv") as dataFile, open("brcaTotalDataTsgTidy.tsv", 'w') as outputFile:
    lineCount = 0
    header = ""
    for line in dataFile:
        if lineCount == 0:
            lineCount += 1
            line = line.strip("\n").split("\t")
            header = line
            outputFile.write("patient\tgene\texpression\n")
        else:
            line = line.strip("\n").split("\t")
            patient = line[0]
            patient = re.sub(r"-\d\d\w-\w\d\d\w-\d\d", r"", patient)
            for i in range(1, len(line)):
                outputFile.write(patient + "\t" + header[i] + "\t" + line[i] + "\n")
            
```

### This python script is used to make the kircTotalDataTsg dataset tidy and remove last portion of patient IDs
```{python}
import re
with open("kircTotalDataTsg.tsv") as dataFile, open("kircTotalDataTsgTidy.tsv", 'w') as outputFile:
    lineCount = 0
    header = ""
    for line in dataFile:
        if lineCount == 0:
            lineCount += 1
            line = line.strip("\n").split("\t")
            header = line
            outputFile.write("patient\tgene\texpression\n")
        else:
            line = line.strip("\n").split("\t")
            patient = line[0]
            patient = re.sub(r"-\d\d\w-\w\d\d\w-\d\d", r"", patient)
            for i in range(1, len(line)):
                outputFile.write(patient + "\t" + header[i] + "\t" + line[i] + "\n")
            
```

### Imports tidy data from previous python scripts
```{r}
brcaTotalDataTsgTidy <- read_tsv("brcaTotalDataTsgTidy.tsv")

kircTotalDataTsgTidy <- read_tsv("kircTotalDataTsgTidy.tsv")

```

### Exports a file with all the patients for a particular type of cancer
```{r}
brcaPatients <- as_data_frame(unique(brcaTotalDataTsgTidy$patient))
brcaPatients <- rename(brcaPatients, patient = value)
write_tsv(brcaPatients, "brcaPatients.tsv")

kircPatients <- as_data_frame(unique(kircTotalDataTsgTidy$patient))
kircPatients <- rename(kircPatients, patient = value)
write_tsv(kircPatients, "kircPatients.tsv")
```

### Removes unessesary files
```{bash}
rm brcaTotalDataTsg.tsv kircTotalDataTsg.tsv brcaTotalDataTsgTidy.tsv kircTotalDataTsgTidy.tsv
``` 

### Combine brca expression and genotype data
```{r}
brcaGenotypeData <- read_tsv("brca_gemini_query_tidy_simplifiedGenotype.vcf")

brcaCombinedData <- inner_join(brcaTotalDataTsgTidy, brcaGenotypeData)

# Create a dataset with depths 20 or above
brcaCombinedDataDepth20Above <- filter(brcaCombinedData, depth >= 20)
```

### BRCA Plots using dataset with depths 20 or above
```{r}

# Impact Severity
ggplot(brcaCombinedDataDepth20Above) +
  geom_boxplot(mapping = aes(x = gene, y = expression, color = impact_severity)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("BRCA: Impact Severity Effect on Gene Expression in Normal Tissue")

ggsave("brcaExpressionByImpactSeverity.png", width = 20, height = 10)

# Genotype
ggplot(brcaCombinedDataDepth20Above) +
  geom_boxplot(mapping = aes(x = gene, y = expression, color = genotype)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("BRCA: Genotype Effect on Gene Expression in Normal Tissue")

ggsave("brcaExpressionByGenotype.png", width = 20, height = 10)

# Mutations per gene
brcaMutationsPerGene <- group_by(brcaCombinedDataDepth20Above, gene, patient) %>% summarise(numberVariants = n(), avgExpression = mean(expression))

ggplot(brcaMutationsPerGene) +
  geom_raster(mapping = aes(x = gene, fill = avgExpression, y = numberVariants)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_fill_gradientn(colours=c("#56B4E9","#D55E00")) +
  labs( x = "Gene", y = "Number of Variants per Patient", fill = "Average Gene Expression", title ="BRCA: The Effect of the Number of Variants on Gene Expression in Normal Tissue")

ggsave("brcaMutationsPerGene.png")
```


### Combine KIRC expression and genotype data
```{r}
kircGenotypeData <- read_tsv("kirc_gemini_query_tidy_simplifiedGenotype.vcf")

kircCombinedData <- inner_join(kircTotalDataTsgTidy, kircGenotypeData)

# Create a dataset with depths 20 or above
kircCombinedDataDepth20Above <- filter(kircCombinedData, depth >= 20)
```

```{r}
# Impact Severity
#ggplot(kircCombinedDataDepth20Above) +
  #geom_boxplot(mapping = aes(x = gene, y = expression, color = impact_severity)) +
  #theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  #ggtitle("KIRC: Impact Severity on Gene Expression in Normal Tissue")

#ggsave("kircExpressionByImpactSeverity.png", width = 20, height = 10)

# Genotype
ggplot(kircCombinedDataDepth20Above) +
  geom_boxplot(mapping = aes(x = gene, y = expression, color = genotype)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("KIRC: Genotype Effect on Gene Expression in Normal Tissue")

ggsave("kircExpressionByGenotype.png", width = 20, height = 10)

# Mutations per gene
#kircMutationsPerGene <- group_by(kircCombinedDataDepth20Above, gene, patient) %>% summarise(numberVariants = n(), avgExpression = mean(expression))

#ggplot(kircMutationsPerGene) +
  #geom_raster(mapping = aes(x = gene, fill = avgExpression, y = numberVariants)) +
  #theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  #scale_fill_gradientn(colours=c("#56B4E9","#D55E00")) +
  #labs( x = "Gene", y = "Number of Variants per Patient", fill = "Average Gene Expression", title ="KIRC: The Effect of the Number of Variants on Gene Expression in Normal Tissue")

#ggsave("kircMutationsPerGene.png")
```

### BRCA: Create individual plots for each gene, looking at effect of genotype on exprssion. Expression has been log transformed
```{r}
for (i in vectorTSG) {
  temp <- filter(brcaCombinedDataDepth20Above, gene == i)
  (ggplot(temp) +
  geom_boxplot(mapping = aes(x = gene, y = log(expression), color = genotype)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("BRCA: Genotype Effect on Gene Expression in Normal Tissue"))
  
  file <- paste("brca_individual_gene_plots/", "brca", i, ".png", sep = "")
  ggsave(file)
}

```

### KIRC: Create individual plots for each gene, looking at effect of genotype on exprssion. Expression has been log transformed
```{r}
for (i in vectorTSG) {
  temp <- filter(kircCombinedDataDepth20Above, gene == i)
  (ggplot(temp) +
  geom_boxplot(mapping = aes(x = gene, y = log(expression), color = genotype)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("KIRC: Genotype Effect on Gene Expression in Normal Tissue"))
  
  file <- paste("kirc_individual_gene_plots/", "kirc", i, ".png", sep = "")
  ggsave(file)
}

```

### To be used when genotype data is brought in to R to make paitent names shorter.
```{python}
#with open("brcaPatients.tsv") as dataFile, open("brcaPatientIDsModified.tsv", 'w') as writeFile:
  #for line in dataFile:
    #if "patient" in line:
      #writeFile.write(line)
    #else:
      #line = line.split("-")
      #newPatientID = "-".join(line[0:4])
      #writeFile.write(newPatientID + '\n')
```

```{python}
#with open("kircPatients.tsv") as dataFile, open("kircPatientIDsModified.tsv", 'w') as writeFile:
  #for line in dataFile:
    #if "patient" in line:
      #writeFile.write(line)
    #else:
      #line = line.split("-")
      #newPatientID = "-".join(line[0:4])
      #writeFile.write(newPatientID + '\n')
```


```{r}
#write_tsv(brcaTotalDataTsgTidy, "brcaTotaDataTsgTidy_fixPatients.tsv")
```

```{python}
#import re
#with open("brcaTotaDataTsgTidy_fixPatients.tsv") as dataFile, open("brcaTotaDataTsgTidy_fixedPatients.tsv", 'w') as writeFile:
    #lineCount = 0
    #for line in dataFile:
        #if lineCount == 0:
            #writeFile.write(line)
            #lineCount += 1
        #else:
            #line = re.sub(r"-\d\d\w-\w\d\d\w-\d\d", r"", line)
            #writeFile.write(line)
```

###Boxplot n genes with highest or lowest standard deviation
```{r}
#Function to plot top or bottom n genes with highest or lowest standard deviation, default is top 10 genes
#boxPlotGenesBySd <- function(tidyDataSet, tsgSummary, howManyGenes = 10){
  #Determine which genes have highest standard deviation
  #genesWithHighestStDev <- top_n(tsgSummary, n = howManyGenes, stDev) %>% pull(gene)
  #print(filter(tidyDataSet, gene %in% genesWithHighestStDev) %>% ggplot(aes(gene, expression)) + geom_boxplot() 
        #+ theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)))
#}

#Boxplot of top 5 genes BRCA
#boxPlotGenesBySd(brcaTotalDataTsgTidy, brcaTotalDataTsgSummary, 5)

#Use to save plot as png
#ggsave("brcaBoxPlotTop5.png")

#Boxplot of top 10 genes BRCA
#boxPlotGenesBySd(brcaTotalDataTsgTidy, brcaTotalDataTsgSummary, 10)

#Use to save plot as png
#ggsave("brcaBoxPlotTop10.png")

#Boxplot of top 5 genes KIRC
#boxPlotGenesBySd(kircTotalDataTsgTidy, kircTotalDataTsgSummary, 5)

#Use to save plot as png
#ggsave("kircBoxPlotTop5.png")

#Boxplot of top 10 genes BRCA
#boxPlotGenesBySd(kircTotalDataTsgTidy, kircTotalDataTsgSummary, 10)

#Use to save plot as png
#ggsave("kircBoxPlotTop10.png")
```

### Summary of gene expression that includes median, mean, min value, max value, standard deviation (`stDev`), arranged in decending order by standard deviation
```{r}
#Breast Cancer Patients
#brcaTotalDataTsgSummary <- group_by(brcaTotalDataTsgTidy, gene) %>% summarize(median = median(expression), mean = mean(expression), min = min(expression), max = max(expression), stDev = sd(expression)) %>% arrange(desc(stDev))

#Kidney Renal Clear Cell Carcinoma Patients
#kircTotalDataTsgSummary <- group_by(kircTotalDataTsgTidy, gene) %>% summarize(median = median(expression), mean = mean(expression), min = min(expression), max = max(expression), stDev = sd(expression)) %>% arrange(desc(stDev))
```